---
title: How to Plot a Network's Degree Distribution Using igraph in R
author: Hongtao Hao
date: '2020-08-19'
slug: plot-degree-distribution-using-igraph
output:
  blogdown::html_page:
    toc: yes
    toc_depth: 2
---
******
The R package `igraph` is very useful. However, its `degree()` and `degree_distribution()` function is not so straightforward. Also, it might be challenging to plot degree distribution. 

## Getting a deeper understanding of degree() and degree_distribution()

To explore these two function, let's first draw a simple network. 

```{r echo=TRUE, results='hide', error=FALSE, warning=FALSE, message=FALSE}
library(igraph)
```

```{r simpleNetwork, echo=TRUE, fig.align='center', fig.cap="A simple network, g1"}
g1 <- graph( edges=c(1,2, 1,3, 2,3, 3,4), n=4, directed=FALSE)
plot(g1)
```

Okay, now let's have a look at its nodes' degrees and its degree distribution.

```{r}
deg_1 <- degree(g1)
deg_1
```

But does this result mean? It means that the degree of node1 is $2$, the degree of node2 is $2$, the degree of node3 is $3$, and the degree of node4 is $1$.

Okay. Let's have a look at `g1`'s degree distribution:

```{r}
degree_distribution(g1)
```

What does this result mean then? Does it stand for the value of node $1-4$. No. It means the **relative frequency** of all the degrees, i.e., $1, 2, 3$, as is shown in the result of `degree(g1)`. Wait, but there are four values in the result above, $0.00, 0.25, 0.50, 0.25$. Well, this is because the first value is for degree $0$. 

Conclusion: `degree()` will tell us the degree of all the nodes. `degree_distribution()` will tell us the **relative frequency** of all the degrees. 

Then, we may wonder, what if we don't have all the degrees from the minimum to maximum. For example, if we only have degrees of $1,2,4,9$, will the `degree_distribution()` function show us the relative frequency of degree $5$ (which should be zero)?

Let's try.

```{r anotherSimpleNetwork, echo=TRUE, fig.align='center', fig.cap="Another simple network, g2"}
set.seed(1234)
g2 <- graph( edges=c(1,2, 1,3, 2,3, 3,4, 4,5, 4,6, 4,7, 
                     4,8, 4,9, 4,10, 3,11, 2,12,
                     4,13, 4,14, 2,15, 6,12, 7,10), 
             n=15, directed=FALSE)
plot(g2)
```

```{r}
deg_2 <- degree(g2)
deg_2
```

```{r}
degree_distribution(g2)
```

As we can see above, yes, **it will**. It displays relative frequencies for all possible degrees from $0$ to $9$. 

## Plotting the distribution
Let's plot it. 

First, you might be thinking of histogram:

```{r}
hist(degree(g1),
     xlab = "k",
     ylab = "Frequency",
     main = "Histogram of g1's nodes degrees, without adjusting breaks and ylim",
     col = "skyblue")
```

However,as we can ses above, it doesn't make much sense because all of the degrees are discrete numbers. Let's try bar plot instead. 

```{r}
t1 <- table (deg_1)
sum(t1)
relafreq_1 <- t1/sum(t1)
barplot(relafreq_1, xlab = "k", ylab = "Relative frequencies", 
        main = "Degree distribution of g1, without adjusting ylim",
        col = "orange")
```

It works. 

Let's plot `g2` as well.

```{r g2BarChart, fig.align="center", fig.cap="bar chart for g2 without adusting ylim"}
t2 <- table (deg_2)
sum(t2)
relafreq_2 <- t2/sum(t2)
barplot(relafreq_2, xlab = "k", ylab = "Relative frequencies", 
        main ="Degree distribution of g2, without adjusting ylim",
        col = "orange")
```

We find that the Y-axis scale is too short for our values. I found a solution [here](https://stackoverflow.com/a/50193822/13716814). Before looking at the solution, let's first try solving it by ourselves. 

What's the problem here? The problem is that the Y-axis scale is too short. To adjust Y-axis scale, let's try `ylim` first. 

What do we need to set `ylim`? As [sebkopf](https://stackoverflow.com/users/3239578/sebkopf) noted [here](https://stackoverflow.com/a/23051088/13716814) on [stackoverflow](https://stackoverflow.com/), we simply need to set the minimum and maximum values using, for example, `c()`.  

Naturally, we may think of `max()`. Let's try:

```{r}
barplot(relafreq_2, xlab="k", ylab = "Relative frequencies", 
        main="Degree distribution of g2, adjusting ylim using `c(0,max())`",
        col = "orange",
        ylim = c(0, max(relafreq_2)))
```

The Y-axis scale is still too short. 

We may use `max(t2/sum(t2))+0.1`. It of course will be fine here, setting the Y-axis scale to be $[0,0.5]$:

```{r}
barplot(relafreq_2, xlab="k", ylab = "Relative frequencies", 
        main="Degree distribution of g2, adjusting ylim using `c(0,max()+0.1)`",
        col = "orange",
        ylim = c(0, max(relafreq_2+0.1)))
```

However, this method may not be applicable to other situations. Is there a better way? Yes, let's try the `pretty()` function mentioned above. Read [here](https://stat.ethz.ch/R-manual/R-devel/library/base/html/pretty.html) and [here](https://statisticsglobe.com/pretty-r-function) to understand this function. 

Basically, `pretty()` function will return a sequence of values that are equally spaced and cover the minimum and maximum values of our input. For example:

```{r}
a <- c(0,1,2,3,0.5,1.2,0.001)
pretty(a)
c <- 0.1:0.5
pretty(c)
d<- 0:10
pretty(d)
```

Be aware that `pretty()` won't be able to process a list:

```{r eval=FALSE}
b <- list(0,1,2,3,0.5,1.2, 0.001)
pretty(b)
# There will be an error: 
# Error in min(x) : invalid 'type' (list) of argument
```

Then, what will happen if we use `pretty(t2/sum(t2))`?

```{r}
pretty(t2/sum(t2))
```

It works, right? However, for `ylim`, we need a min and max. How to get them? We can use the [`range()` function](https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/range), which will return [a vector](https://rstudio-education.github.io/hopr/r-objects.html#atomic-vectors) that contains the min and max of our input. 

```{r}
barplot(relafreq_2, xlab="k", ylab = "Relative frequencies", 
        main="Degree distribution of g2, adjusting ylim using `range(pretty())`",
        col = "orange",
        ylim = range(pretty(c(0,relafreq_2))))
```

Some of you may wonder, why couldn't we just use `range(pretty(relafreq_2))`. Well, **we could**. The thing is, if all degrees appeare at least once, then using `range(pretty(relafreq_2))` rather than `range(pretty(c(0,relafreq_2)))` means that the start of `ylim` won't be $0$, I guess. So it's safer to use `range(pretty(c(0,relafreq_2)))`.

## Going back to histogram
In the [igraph tutorial](https://kateto.net/networks-r-igraph) by [Dr. Katya Ognyanova](https://kateto.net/), I saw that she showed the node degrees using histograms. 

On second thought, I realized it's not necessarily a bad idea to show node degrees via a histogram. After all, when the size of the network is huge and the node degrees vary greatly, a histogram isn't that much different from a bar plot. 

Why do I say that they are not that different? I'll explain it using an example of a random graph. 

```{r randomGraph, fig.align="center", fig.cap="A random graph"}
set.seed(1234)
random_graph <- sample_gnp(1000, 1/1000, directed = FALSE, loops = FALSE)
plot(random_graph,
     main = "Plotting random_graph") # It's not very pretty...
```

Let's draw its nodes degree 

```{r rnd-g-hist, fig.cap="Histogram of random_graph's nodes degrees, without adjustments"}
deg_random <- degree(random_graph)
hist(deg_random, 
     main = "Hist of nodes degree for rnd_g, without adjustments",
     xlab = "k",
     col = "skyblue")
```

`r blogdown::shortcode("reminder")`
The above figure has two problems: X-axis and Y-axis. We'll solve them later.
`r blogdown::shortcode("end")`

Then, let's compare the above histogram with the corresponding bar chart:

```{r}
t_random <- table(deg_random)
relafreq_random <- t_random/sum(t_random)
barplot(relafreq_random, xlab = "k",
        ylab = "Relative frequencies",
        main = "Degree distribution of random_graph, with ylim adjusted",
        ylim = range(pretty(c(0,relafreq_random))),
        col = "orange")
```

They are not that different, right? Except for the fact that the histogram captures the absolute frequencies whereas the bar chart displays relative ones. 

### Deciding on X-axis scale

When deciding on the X-axis scale of the histogram, [Dr. Katya Ognyanova](https://kateto.net/) in [one of her posts](https://kateto.net/networks-r-igraph) used `vcount()-1`. This worked in the example she provided but I don't think it is the best option. For example, in the above random graph we have, there are $1,000$ vertices, but the maximum degree is only $5$. Therefore, we cannot use `vcount()-1`, which is $999$ in our case, to denote the maximum on the `X-axis`:

```{r}
max(deg_random)
```

```{r fig.cap="`vcount()-1` doesn't fit quite well here"}
hist(deg_random, breaks = 0:vcount(random_graph)-1, 
     main = "Hist of nodes degree for rnd_g, using `vcount()-1`",
     xlab = "k", 
     col = "skyblue",
     ylim = range(pretty(c(0,table(deg_random))))
     )
```

As we can see above, the figure doesn't show the nodes degrees in detail because the X-axis is too large. A better alternative might be `max(degree(G))+1` where `G` is the graph you are interested in:

```{r}
hist(deg_random, breaks = 0:(max(deg_random)+1), 
     main = "Hist of nodes degree for rnd_g, adjusting breaks using `seq()`",
     xlab = "k", 
     col = "skyblue"
     )
```

### Adjusting a histogram's breaks
As we can see above, the range of [0,1] shows the frequencies of $k=0$ and $k=1$ combined:

```{r}
t_random
```

This problem also occurs in Figure \@ref(fig:rnd-g-hist)

How to solve this problem? 

We can simply change the breaks. Set it as `breaks = seq(0,(max(deg_random)+1), by=0.5)` so that there [0,1] will be divided into two parts. I learned this trick from [here](https://stackoverflow.com/a/14585847/13716814).

```{r}
hist(deg_random, breaks = seq(0,(max(deg_random)+1), by=0.5), 
     main = "Hist of nodes degree for rnd_g, adjusting breaks using `seq()`",
     xlab = "k", 
     col = "skyblue"
     )
```

However, it's still a little packed in $[0,1]$. Let's try setting the start of the `seq()` to be $-1$:

```{r}
hist(deg_random, breaks = seq(-1,(max(deg_random)+1), by=0.5), 
     main = "Hist of nodes degree for rnd_g, adjusting breaks using `seq()`",
     xlab = "k", 
     col = "skyblue"
     )
```

Now, it's much clearer. 

### Adjusting a histogram's Y-axis
In the above figures, the Y-axis scale is too short. How to solve this problem?

We can simply use the trick of `range(pretty())` again here:

```{r}
hist(deg_random, breaks = seq(-1,(max(deg_random)+1), by=0.5), 
     main = "Hist of nodes degree for rnd_g, adjusting Y-axis using `range(pretty())`",
     xlab = "k", 
     col = "skyblue",
     ylim = range(pretty(c(0,table(deg_random))))
     )
```

## Conclusion

### How to draw a bar chart to display a network's degree distribution

For bar chart, the only problem we faced lies in the `ylim`. We simply need to use `ylim = range(pretty())`

### How to draw a histogram to display a network's nodes degrees

When drawing a histogram, we had three issues: 

1. Deciding on X-axis scale 
1. Breaks in X-axis
2. Y-axis scale might be too short

To decide on the X-axis scale, we tried `vcount()-1` but the result was not ideal. Then we used `max(degree(G))`, which is better. 

The second problem is how to set the breaks. We first used `breaks = 0:(max(G)+1)` but it will show $k=0$  and $k=1$ combined, which is not what we want. Then, we tried `breaks = seq(0,(max(deg_random)+1), by=0.5)`. It got better but was still not quite ideal. Then we set the start of the `seq()` to be $1$ and the result was good enough. 

The third issue is the short Y-axis scale. We solved the problem using `range(pretty())`. 

## Another option
[Chengjun Wang](https://chengjunwang.com/) [defined](https://chengjunwang.com/web_data_analysis/demo2_simulate_networks/) the function `plot_degree_distribution()` and `fit_power_law()`, which I found to be very useful. 