---
title: Reading Online and Local Data Files in Rmarkdown, Bookdown, and Blogdown
author: Hongtao Hao
date: '2020-08-24'
slug: reading-online-and-local-files-in-rmarkdown
categories: []
tags: []
output:
  blogdown::html_page:
    toc: yes
---
******
[Here](https://blogdown-demo.rbind.io/2018/02/27/r-file-paths/) is a tutorial on how to read local data files in Rmarkdown. I found it very helpful. Nonetheless, I decided to write about this topic myself because the post is more complicated than it should be. 

We can read data files from two sources: online, or local. 

# Reading Online Files

Reading online files is easy:

```{r warning=FALSE, message=FALSE, eval=FALSE}
library(readr)
hair <- read_csv("https://hongtaoh.com/files/HGN.csv")
```

# Reading Local Files

To read local files in Rmarkdown, you can either specify the exact paths, or use the package of `here`. One thing worth noticing is that `R` can read any local files, either they are in the same root directory as your `Rmarkdown` file or not, if you specify the exact paths. However, if you use the package of `here`, the data file to be read must be in the same directory as your `Rmarkdown` file. 

## Specifying Paths

```{r eval=FALSE}
hair <- read_csv("/path/to/your/file/filename.csv")
```