---
title: Using Rmarkdown in Hugo
author: Hongtao Hao
date: '2020-01-28'
slug: using-markdown-in-hugo-post
categories: []
draft: true
tags: []
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Exercise 4.1 
```{r}
df <- HairEyeColor # Changing the name of HairEyecolor
show (df) # displaying the dataframe
e.h.freq <- apply (df, c('Eye','Hair'), sum) # To get a table where 'Eye color' is the row and 'Hair color' is the column. This line of code will sum across sex. 
e.h.prop <- e.h.freq / sum (e.h.freq) # Generating joint proportions in each cell.
show (round (e.h.prop, 2)) # Round off the values in e.h.prop to the 2nd number of decimal places and then display the dataframe. 
h.freq <- apply (df, c('Hair'), sum) # This will sum across sex and eye, and display the number of people having a specific hair color. 
h.prop <- h.freq / sum (h.freq) # This will display the marginal propabilities (Hair)
show ( round(h.prop,2) ) # Round off and display the dataframe.
e.freq <- apply (df, c('Eye'), sum) # This will sum across sex and hair, and display the number of people having a specific eye color.
e.prop <- e.freq /sum (e.freq) # This will display the marginal probabilities (Eye)
show (round(e.prop, 2) ) # Round off and display the dataframe. 
e.h.prop ['Blue', ] / e.prop ['Blue'] # This is p(h|e=blue), i.e., the probabilities of the hair colors given that eye color is blue. 
# To get the probabilities of hair colors given Brown eyes:
e.h.prop ['Brown',] / e.prop ['Brown']
# To get the probabilities of eye colors given Brown hair:
e.h.prop [, 'Brown'] / h.prop ['Brown']

```

## Exercise 4.2
```{r}
set.seed(47405)
N <- 500
pHeads <- 0.8
flipSequence <- sample (x = c(0,1), prob = c(1-pHeads,pHeads), size = N, replace = T)
num.heads <- cumsum (flipSequence)
num.flips <- c (1:N)
prop.heads <- num.heads / num.flips
plot (num.flips, prop.heads, type = 'o', log='x', 
      col="skyblue", xlim = c(1,N), ylim = c(0.0,1.0), cex.axis = 1.5,
      xlab = "Flip Number", ylab = "Proportion Heads", cex.lab = 1.5,
      main = "Running Proportion of Heads", cex.main=1.5)
abline( h=pHeads, lty="dotted")
flipLetters <- paste (c("T", "H")[flipSequence[1:10]+1],collapse = "")
displayString <- paste0( "Flip Sequence =" , flipLetters , "...")
text (N, .5, displayString, adj = c(1,0.5), cex = 1.3)
text (N, .4, paste("End Proportion = ", prop.heads[N]), adj=c(1,0.5), cex=1.3)
```
## 3
```{r}
set.seed(1)
x <- seq(5, 15, length=1000)
y <- dnorm(x, mean=10, sd=.20)
plot(x, y, type="l", lwd=1, col="red")

```

### 3A
```{r}
set.seed(1)
dnorm(9.9,mean=10, sd=.20)
```

### 3B
```{r}
set.seed(1)
# Generating a random sample of 100,000 values from the normal distribution:
heapOdata <- rnorm (100000, mean = 10.0, sd = .20)
hist(heapOdata, breaks = 51, col = "orange", main = "rnorm") # Having a quick overview of this set of random numbers.
```

### 3C
```{r}
num.between <- sum( heapOdata >= 9.8 & heapOdata < 10.0 ) # counting how many numbers fall between the interval
prob.mass <- num.between / 100000 # Calculating probability mass
approx.prob <- prob.mass / (10.0-9.8) # Calculating probability density (probability mass / bin width)
approx.prob 
# which is approximately the same as Part A.
```

## 4

According to Baye's rule, among all joint outcomes with "negative results", the proportion of people having the disease is $p (user|-)$. Specifically:

$$p (user|-) = \frac{p (-|user)\cdot p(user)}{p(-)}$$
We should notice that the prior, $p(user)$ is now the posterior computed from the [Table 5.4](https://www.sciencedirect.com/science/article/pii/B9780124058880000052#t0025) in the book. Also notice that,

$$P(-) = \sum_{θ^*} \cdot P (-|θ^∗) \cdot P(θ^*)  $$
Now, I will compute the posterior from [Table 5.4](https://www.sciencedirect.com/science/article/pii/B9780124058880000052#t0025):

$$P (user|+) = \frac {0.99 \cdot 0.001}{0.99 \cdot 0.001 + 0.05 \cdot (1-0.001)}$$
Using R, we know that the result is:

```{r}
posterior <- (0.99*0.001)/(0.99*0.001+0.05*(1-0.001))
posterior
```


\[
\begin{eqnarray}
P (user|-) & = & \frac{P (-|user)\cdot P(user)}{P(-)}\\
\\
& = & \frac{P (-|user)\cdot P(user)}{\sum_{θ^*} \cdot P (-|θ^∗) \cdot P(θ^*)}\\
\\
& = & \frac {0.01 \cdot 0.01943463 }{P(-|user) \cdot P(user) + P(-|non-user) \cdot P (non-user)} \\
\\
& = & \frac {0.01 \cdot 0.01943463 }{0.01 \cdot 0.01943463 + 0.95 \cdot (1-0.01943463)}
\end{eqnarray}
\]

Calculating using R:

```{r}
(0.01*posterior) / {0.01*posterior+(0.95*(1-posterior))}
```

Therefore, the probability that the person has the disease given that the re-test shows "negative" is 0.0002085862.

## 5

### 5A
Test     User   Non-user   Margin 
------- ------ ---------- ------- 
+         99      4995     5094 
-         1       94905    94906 
Margin   100    99,900    100,000

### 5B
The proportion of people who have the disease given that the test result is positive:

```{r}
99/5094
```

### 5C & 5D
```{r}
upper.left <- 10000 * 0.99
upper.left
upper.right <- 9990000 * 0.05
upper.right
lower.left <- 10000 * 0.99 * 0.01
lower.left
lower.right <- 9990000 * 0.05 * 0.95
lower.right
lower.left / (lower.left+lower.right) # The result is the same as the answer to Exercise 5.1. 
```
